---
title: "noaamsdr: Capstone project for Coursera"
author: "Rashid Kazmi"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This is the final project of the Coursera's "Master Software Development in R"
specialization. The objective of this package is provide wrapper functions to
quickly analyse and visualize the National Oceanic and Atmospheric Administration
dataset of significant earthquakes.

Currently the package have the following capabilities:

- Get the NOAA earthquakes dataset
- Clean the data
- Visualize the earthquakes timeline
- Visualize the earthquakes in a map

__NOTE:__ This package heavily relies on the Hadley Wickham's [tidyverse](tidyverse).

### Get data
This package will download and put the data in a persistent directory unless you
specify another path. Just write 

```{r}
library(tidyverse)
library(lubridate)
library(noaamsdr)

eq_get_data() %>% glimpse
```

and you will have the complete raw NOAA dataset in a data.frame object with all
variables imported as character. The type handling is done in the cleaning process.
This function depends on a the low level functions `eq_download_data` and `eq_file_path`.
The `eq_download_data` download the NOAA earthquakes dataset in the specified
directory. The `eq_file_path` returns a valid absolute path to the dataset, if
the file to not exist in the passed directory, it downloads it. If the destdir is
not specified, a persistent directory will be created. This is the default and
prefered behaviur. The persistent directory is OS dependent. If you want to update
the dataset pass the update = TRUE param.

### Clean data
By default, all variables are imported as character. This is because we want to
make a more robust cleaning of the data. Again, just pass the dataframe to 
the cleaning data and you will have a clean dataset with the types of each variable
corrected and some new variables that compact information (e.g. DATE column).

```{r}
eq_get_data() %>% eq_clean_data %>% head
```

This function relies on several internal functions, namely:

- numeric_convertible: Check if a data.frame column can be safely converted to
numeric type.
- eq_clean_wrong_parsed_numeric_vectors: Mutate all the columns that can be safely
converted to numeric type.
- eq_clean_dates: Unite YEAR, MONTH and DAY columns into a single "DATE" column.
- eq_clean_coordinates: Remove invalid coordinate rows.
- eq_clean_location: Removes the country from the LOCATION variable.

All this functions are implemeted for interal usage.


### Visualize a timeline of the dataset
It could be useful to visualize the earthquakes across time and other variables
This is implemented in a couple of ggplot2's geoms: `geom_timeline` and `geom_timeline_label`.
The `geom_timeline` works out of the box with the required aes "x". The `geom_timeline_label`
requires three aes: "x", "label" and "by". The "by" aesthetic is alog with the "n_max"
param to show only top n_max labels in the by param.

```{r}
df <- eq_get_data() %>%
  eq_clean_data %>%
  filter(!is.na(EQ_PRIMARY)) %>%
  filter(year(DATE) > 2000) %>%
  filter(COUNTRY %in% c("CHINA", "USA"))

ggplot(df) +
  aes(
    x = DATE,
    y = COUNTRY,
    size = EQ_PRIMARY,
    colour = DEATHS,
    label = LOCATION_NAME,
    by = EQ_PRIMARY
  ) +
  geom_timeline() +
  geom_timeline_label(n_max = 5) +
  theme_timeline
```

The package also provides the `theme_timeline` wrapper for a clean visualization.

### Visualize the earthquakes in a map
A leaflet wrapper is implemented to easily visualize the epicenters:

```{r, eval=F}
eq_get_data() %>%
  eq_clean_data %>%
  filter(COUNTRY == "MEXICO", year(DATE) >= 2000) %>%
  eq_map(annot_col = "DATE")

eq_get_data() %>%
  eq_clean_data %>%
  filter(COUNTRY == "MEXICO", year(DATE) >= 2000) %>%
  mutate(popup_text = eq_create_label(.)) %>%
  eq_map(annot_col = "popup_text")
```


## References

- National Geophysical Data Center / World Data Service (NGDC/WDS): 
  Significant Earthquake Database.
  National Geophysical Data Center, NOAA. doi:10.7289/V5TD9V7K

[NOAA]: https://www.ngdc.noaa.gov/nndc/struts/form?t=101650&s=1&d=1
[tidyverse]: https://github.com/tidyverse/tidyverse
